A multi-species benchmark for training and validating large scale mass spectrometry proteomics machine learning models

### 2024-07-17 doc/nine-species/nine-species.tex
Minor edits

\textit{De novo} sequencing of proteomics tandem mass spectrometry data, in which observed fragmentation spectra are translated into corresponding peptide sequences, has been an open challenge for more than 40 years \cite{sakurai1984paas}.
Recently, as in many other areas of science, considerable progress toward solving this challenge has been made using deep learning, in which multi-layer neural networks with millions of parameters are trained to generate peptide sequences from observed spectra.
The first such deep learning method, DeepNovo \cite{tran2017denovo}, has been followed by at least 22 additional publications (reviewed in \cite{XXX}).

The standard method for evaluating these \textit{de novo} sequencing methods is to use a gold standard produced via database search.
In this approach, mass spectrometry data derived from a single species is searched against the reference proteome for that species, yielding a ranked list of peptide-spectrum matches (PSMs).
Including in the peptide database a collection of reversed or shuffled ``decoy'' peptides provides a rigorous way to set a threshold in this list of PSMs while controlling the false discovery rate (FDR) among the PSMs above the threshold \cite{elias:target}.
The resulting set of high-confidence PSMs can be used either to train or evaluate a \textit{de novo} sequencing model.

@@

Unfortunately, creating a high quality gold standard set of labeled spectra can be tricky.
One challenge is ensuring that the search strategy employs appropriate parameters.
For instance, one widely used benchmark dataset \cite{tran2017denovo} used a search strategy that failed to account for missassigned isotopic peaks during the acquisition stage.
This error led to frequently assigning a deamidation modification, when the observed mass shift was better explained by an isotopic mass shift on the precursor \textit{m/z} \cite{bittremieux:deep}.
A second challenge relates to the notion of train/test ``leakage,'' in which information used to train the model leaks into the evaluation procedure.
In the \textit{de novo} setting, a common mistake is to randomly segregate a given set of labeled spectra into training and test sets, without regard to the associated peptides.
As a result, spectra generated by the same peptide sequence may occur in both the training and test sets.
Such duplicated peptides give an unfair advantage to the sequencing method, and the leakage will be even more useful to parameter-rich methods that are capable of memorizing many features of the training data.

In this work, we revisit the nine-species benchmark dataset that was employed in the first deep learning \textit{de novo} sequencing method, DeepNovo \cite{trans2017denovo}.
This is a widely used dataset, which has been employed for training or evaluation in at least \fixme{XXX} subsequent studies \cite{XXX}.
The setup is quite straightforward.
The authors downloaded nine publicly available datasets, all of which were generated on a Thermo Scientific Q Exactive mass spectrometer, and each of which was carried out in a different species.
Each dataset was searched against the corresponding reference proteome, using a target-decoy strategy to accept a set of PSMs subject to a PSM-level FDR threshold of 1\%.
Because the data are derived from different species, the peptides in each set are largely (but not entirely) disjoint.
To use the benchmark, it is typical to apply a cross-validation strategy, in which a model is trained on eight species and tested on the held-out species, and the procedure is repeated nine different ways.

In developing our Casanovo \textit{de novo} sequencing model, we identified several problems with the nine-species benchmark \cite{yilmaz2023sequence}.
These included the deamidation problem mentioned above, as well as some uncertainty regarding how the FDR was controlled.
Perhaps most importantly, we recognized that a non-negligible proportion of peptides are shared among the different species, with the highest overlap between human and mouse.

>>>>

\textit{De novo} sequencing of proteomics tandem mass spectrometry data, in which observed fragmentation spectra are translated into corresponding peptide sequences, has been an open challenge for more than 40 years \cite{sakurai1984paas}.
Recently, as in many other areas of science, considerable progress toward solving this challenge has been made using deep learning, in which multi-layer neural networks with millions of parameters are trained to generate peptide sequences from observed spectra.
The first such deep learning method, DeepNovo \cite{tran2017denovo}, has been followed by at least 22 additional publications (reviewed in \cite{bittremieux2024deep}).

The standard method for evaluating these \textit{de novo} sequencing methods is to use a gold standard produced via database search.
In this approach, mass spectrometry data derived from a single species is searched against the reference proteome for that species, yielding a ranked list of peptide-spectrum matches (PSMs).
Including in the peptide database a collection of reversed or shuffled ``decoy'' peptides provides a rigorous way to set a threshold in this list of PSMs while controlling the false discovery rate (FDR) among the PSMs above the threshold \cite{elias2007target}.
The resulting set of high-confidence PSMs can be used either to train or evaluate a \textit{de novo} sequencing model.

@@

Unfortunately, creating a high quality gold standard set of labeled spectra can be tricky.
One challenge is ensuring that the search strategy employs appropriate parameters.
For instance, one widely used benchmark dataset \cite{tran2017denovo} used a search strategy that failed to account for missassigned isotopic peaks during the acquisition stage.
This error led to frequently assigning a deamidation modification, when the observed mass shift was better explained by an isotopic mass shift on the precursor \textit{m/z} \cite{bittremieux2024deep}.
A second challenge relates to the notion of train/test ``leakage,'' in which information used to train the model leaks into the evaluation procedure.
In the \textit{de novo} setting, a common mistake is to randomly segregate a given set of labeled spectra into training and test sets, without regard to the associated peptides.
As a result, spectra generated by the same peptide sequence may occur in both the training and test sets.
Such duplicated peptides give an unfair advantage to the sequencing method, and the leakage will be even more useful to parameter-rich methods that are capable of memorizing many features of the training data.

In this work, we revisit the nine-species benchmark dataset that was employed in the first deep learning \textit{de novo} sequencing method, DeepNovo \cite{tran2017denovo}.
This is a widely used dataset, which has been employed for training or evaluation in at least \fixme{XXX} subsequent studies \cite{XXX}.
The setup is quite straightforward.
The authors downloaded nine publicly available datasets, all of which were generated on a Thermo Scientific Q Exactive mass spectrometer, and each of which was carried out in a different species.
Each dataset was searched against the corresponding reference proteome, using a target-decoy strategy to accept a set of PSMs subject to a PSM-level FDR threshold of 1\%.
Because the data are derived from different species, the peptides in each set are largely (but not entirely) disjoint.
To use the benchmark, it is typical to apply a cross-validation strategy, in which a model is trained on eight species and tested on the held-out species, and the procedure is repeated nine different ways.

In developing our Casanovo \textit{de novo} sequencing model, we identified several problems with the nine-species benchmark \cite{yilmaz2024sequence}.
These included the deamidation problem mentioned above, as well as some uncertainty regarding how the FDR was controlled.
Perhaps most importantly, we recognized that a non-negligible proportion of peptides are shared among the different species, with the highest overlap between human and mouse.
